---
title:  "Coursera, Reproductible Research, Assignment: Course Project 1"
author: "Gauthier le Courtois du Manoir"
date: "10 avril 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a work for Coursera "Reproducible Research" course, Week 1  

## Download & preprocessing the data
```{r downloaddata}
library(xtable)
library(dplyr)
library(lattice)
setwd("C:/Users/Gauthier/Documents/R Working Directory/Reproducible Research/")
if(!file.exists("./data")){dir.create("./data/")}
url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
sourcefile<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile<-"./data/activity.zip"
csvfile<-"./data/activity.csv"
download.file(sourcefile, destfile)
unzip(destfile,exdir = "data")
activity<-read.csv(file = csvfile, na.strings = "NA" )
head(activity)
```
##Mean total number of steps taken per day
```{r}
sum<-data.frame(tapply(activity$steps, activity$date, sum, na.rm=TRUE))
Totstepperday<-cbind(rownames(sum),sum)
rownames(Totstepperday)=c(1:length(unique(activity$date)))
colnames(Totstepperday)=c("Date","Tot_num_step")
mean_all_days<-rep(mean(Totstepperday$Tot_num_step),length(unique(activity$date)))
names(mean_all_days)=c("Mean")
Totstepperday<-cbind(Totstepperday,mean_all_days)

#head(Totstepperday)

hist(Totstepperday$Tot_num_step,xlab="Total number of Steps per day",ylab="Frequency in days",ylim=c(0,30),main="Total number of steps taken per day",col="lightblue", breaks=10)
summary(Totstepperday$Tot_num_step)
abline(v=10000,col="red")
legend("topleft", legend ="mean",lty=1, col="red")
```
  
For the total number of step per day, the mean is **`r mean(Totstepperday$Tot_num_step)`** and the median is **`r median(Totstepperday$Tot_num_step)`**
  

## Average daily activity pattern
```{r}

tmp<-data.frame(tapply(activity$steps, activity$interval, sum, na.rm=TRUE))
Totstepperintervals<-cbind(as.integer(rownames(tmp)),tmp)
colnames(Totstepperintervals)=c("Interval","Tot_num_step")
rownames(Totstepperintervals)=c(1:length(unique(Totstepperintervals$Interval)))

plot(Totstepperintervals$Interval,Totstepperintervals$Tot_num_step, type="l",col="blue",xlab="Intervals",ylab="Total number of steps taken per interval",main="Average number of steps taken")

abline(h=mean(Totstepperintervals$Tot_num_step, na.rm=T),type="l",col="red")
points(subset(Totstepperintervals,Tot_num_step==max(Totstepperintervals$Tot_num_step,na.rm=T))$Interval,max(Totstepperintervals$Tot_num_step,na.rm=T),col="red")
legend("topleft", legend ="mean",lty=1, col="red")
```
  
The interval **`r subset(Totstepperintervals,Tot_num_step==max(Totstepperintervals$Tot_num_step,na.rm=T))$Interval`** has the maximum number of step which is **`r max(Totstepperintervals$Tot_num_step,na.rm=T)`**.
  
## Imputing missing values
Number of missing values in the activity dataset is **`r sum(is.na(activity$steps))`**.

We replace the NA value for each day by the mean value for the day.

```{r Nmissing_value}
#We replace NA value by the mean of steps per day
mean_by_day<-data.frame(tapply(activity$steps, activity$date, mean, na.rm=TRUE))
na_mean<-activity
na_mean$date<-as.character(na_mean$date)
#Date which have NA as value of steps
a<-0
for (i in 1:nrow(na_mean)){
  if(is.na(na_mean$steps[i])){
      if(is.na(mean_by_day[na_mean$date[i],])==TRUE){
          na_mean$steps[i]<-0
      }else{
          na_mean$steps[i]<-mean_by_day[na_mean$date[i],]
      }
    a<-sum(a,mean_by_day[na_mean$date[i],], na.rm=T)
    #print(mean_by_day[na_mean$date[i],])
   # na_mean$steps[i]<-mean_by_day[na_mean$date[i],]
  }
}
```
  
All the missing value are for all the day. So mean, median and sum of steps haven't changed.There is no impact with using mean of median to replace NA value.

```{r Creating_plot}
#Create a dataset to have the total number of step per day with this new dataset
sum_na_mean<-data.frame(tapply(na_mean$steps, na_mean$date, sum, na.rm=TRUE))
Totstepperday_na_mean<-cbind(rownames(sum_na_mean),sum_na_mean)
rownames(Totstepperday_na_mean)=c(1:length(unique(activity$date)))
colnames(Totstepperday_na_mean)=c("Date","Tot_num_step")
mean_all_days_na_mean<-rep(mean(Totstepperday_na_mean$Tot_num_step),length(unique(activity$date)))
names(mean_all_days_na_mean)=c("Mean")
Totstepperday_na_mean<-cbind(Totstepperday_na_mean,mean_all_days_na_mean)

#creating the plot
hist(Totstepperday_na_mean$Tot_num_step,xlab="Total number of Steps per day after transforming NA value by the mean of day",ylab="Frequency in days",ylim=c(0,30),main="Total number of steps taken per day",col="lightblue", breaks=10)
#summary(Totstepperday_na_mean$Tot_num_step_na_mean)
abline(v=mean(Totstepperday_na_mean$Tot_num_step),col="red")
legend("topleft", legend ="mean",lty=1, col="red")
```
  
For the total number of step per day, after replacing NA value by the mean of step of the day, the mean is **`r mean(Totstepperday_na_mean$Tot_num_step)`** and the median is **`r median(Totstepperday_na_mean$Tot_num_step)`**
  
As said before, there is no impact in replacing NA value by the mean of step of the day.

##Are there differences in activity patterns between weekdays and weekends? :

```{r}
#We create a set of data with weekday/weekend factor variable
Sys.setlocale("LC_TIME", "English") # to have weekday in english
na_mean2<-cbind(na_mean,as.character(weekdays(as.Date(na_mean$date))))
names(na_mean2)<-c("steps","date","interval","day")

for (i in 1:nrow(na_mean2)){
  if(na_mean2$day[i]=="Monday" | na_mean2$day[i]=="Tuesday" | na_mean2$day[i]=="Wednesday" | na_mean2$day[i]=="Thursday" | na_mean2$day[i]=="Friday"){
    na_mean2$typeday[i]<-"Weekday"
  }else {
        if(na_mean2$day[i]=="Saturday" | na_mean2$day[i]=="Sunday"){ 
            na_mean2$typeday[i]<-"Weekend"
        }
  }
}
na_mean2$typeday<-factor(na_mean2$typeday)

#We plot the data
plot_na_mean <- na_mean2 %>% 
        select(steps, interval, typeday) %>%
        group_by(typeday,interval ) %>%
        summarise(mean(steps)) 
  names(plot_na_mean)<-c("typeday","interval","mean_steps")

  xyplot(mean_steps~interval| typeday, type="l",data=plot_na_mean)  
  
```
  
As we can see on the plots,   
- During Weekday, the average number of steps begins ealier than in Weekend and ends also earlier then in Weekend  
- The average of steps is higher in Weekend than in Weekday  
- There is a pic of activity in Weekday between interval 500 and 100  
